Supporting Information: Thermal sensitivity across forest vertical profiles: patterns, mechanisms, and ecological implications

### 2021-10-01 vertical-thermal-review.Rmd
"finished" warming section
... including addition of some somewhat wild speculation. :-)

@NidhiVinod , see what you think.

In turn, the drought sensitivity of woody growth tends to be greater in canopy trees. 
In the field of dendrochronology, it is generally accepted that tree ring records of large, exposed trees are best suited for climate reconstructions because their annual growth displays the greatest sensitivity to interannual variation in climate [@frittsTreeRingsClimate1976]. 
However, only a relatively limited number of studies have directly examined drought- or temperature-sensitivities as a function of tree size.
These have generally found greater sensitivity to low precipitation or seasonally high temperatures among larger, more exposed trees [Fig. `r fig_treerings`, @trouillierSizeMattersComparison2018; @gillerotTreeNeighbourhoodDiversity2020; @mcgregorTreeHeightLeaf2021], corroborating evidence from globally disurbed forest censuses that larger trees tend to undergo larger growth declines during drought [@bennettLargerTreesSuffer2015]. 
In addition to lower drought resistance of growth, larger trees frequently exhibit lower ability to recover from stress [i.e., resilience, @mcgregorTreeHeightLeaf2021; @gillerotTreeNeighbourhoodDiversity2020], and greater increases in mortality [@bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019].
Mechanistically, this is almost certainly driven in part by the fact that larger trees have their crowns in a microenvironment that is more challenging during drought [Figs. `r fig_schematic` - `r fig_leaf_T`, @scharnweberConfessionsSolitaryOaks2019], yet there is also reason to believe that height itself provides disadvantages [@olsonPlantHeightHydraulic2018; @couvreurWaterTransportThrough2018].  
Indeed, despite the potential for shorter trees in open forests to experience greater environmental stress [@curtis_intracanopy_2019], greater drought sensitivity of larger trees has been observed in open as well as closed-canopy forests [@bennettLargerTreesSuffer2015; @anderson-teixeiraCarbonCyclingMature2021], supporting the latter concept.
The relative importance of exposure versus height in shaping drought sensitivity remains to be disentangled.

@@

To the extent that warming is coupled to drought, we expect that the largest trees will usually be hardest-hit, at least in part because their crowns are positioned in a more challenging microenvironment (**FIGS, REFS**).
*Unknown: Is there a difference between soil  drought and atmospheric drought, with the most extreme case being a heat wave under well-watered conditions?*

@@

Thus, in synthesis, warming on its own...
*In the understory, warming is disproportionately affecting the less thermally-adapted plant species, resulting in associated plant community thermophilization [@zellwegerForestMicroclimateDynamics2020; @duqueThermophilizationAdultJuvenile2015; @greiserHidingClimateCharacterizing2019].*

@@

>>>>

In turn, the drought sensitivity of woody growth tends to be greater in canopy trees. 
In the field of dendrochronology, it is generally accepted that tree ring records of large, exposed trees are best suited for climate reconstructions because their annual growth displays the greatest sensitivity to interannual variation in climate [@frittsTreeRingsClimate1976]. 
However, only a relatively limited number of studies have directly examined drought- or temperature-sensitivities as a function of tree size.
These have generally found greater sensitivity to low precipitation or seasonally high temperatures among larger, more exposed trees [Fig. `r fig_treerings`, @trouillierSizeMattersComparison2018; @gillerotTreeNeighbourhoodDiversity2020; @mcgregorTreeHeightLeaf2021], corroborating evidence from globally disurbed forest censuses that larger trees tend to undergo larger growth declines during drought [@bennettLargerTreesSuffer2015]. 
In addition to lower drought resistance of growth, larger trees frequently exhibit lower ability to recover from stress [i.e., resilience, @mcgregorTreeHeightLeaf2021; @gillerotTreeNeighbourhoodDiversity2020], and greater increases in mortality [@bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019].
Mechanistically, this is almost certainly driven in part by the fact that larger trees have their crowns in a microenvironment that is more challenging during drought [Figs. `r fig_schematic` - `r fig_leaf_T`, @scharnweberConfessionsSolitaryOaks2019], yet there is also reason to believe that height itself provides disadvantages [@olsonPlantHeightHydraulic2018; @couvreurWaterTransportThrough2018].  
Indeed, despite the potential for shorter trees in open forests to experience greater environmental stress [@curtis_intracanopy_2019], greater drought sensitivity of larger trees has been observed in open as well as closed-canopy forests [@bennettLargerTreesSuffer2015; @anderson-teixeiraJointEffectsClimate2021], *although there is also evidence that short trees in young stands are more drought-sensitive than taller trees in mature forests [**REFS**]*.
The relative importance of exposure versus height in shaping drought sensitivity remains to be disentangled.

@@

To the extent that warming is coupled to drought, we expect that the largest trees will usually be hardest-hit, particularly in cases where the drought is severe, and that this will occur at least in part because their crowns are positioned in a more challenging microenvironment [section 5.1, Figs. `r fig_schematic`- `r fig_treerings`, e.g., @bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019; @anderson-teixeiraJointEffectsClimate2021].
Specifically, warming will disproportionately stress tall canopy trees when $T_{air}$ and accompanying VPD rise enough that sun-exposed leaves cannot maintain both hydraulic safety and the transpirational cooling required to keep $T_{leaf}$ below damaging thresholds.
In contrast, understory trees will be more sheltered during droughts and heat waves, and in some settings may benefit from increased light availability [@bennettLargerTreesSuffer2015].
An interesting open question is to what extent these patterns vary with the nature of the drought, including the relative contributions to stress from low soil moisture versus high VPD. 
While the two are coupled over longer time scales [@humphrey_soil_2021; @novickIncreasingImportanceAtmospheric2016], the latter can be intense for short periods even when soil moisture is high (e.g., during a heat wave) and exerts a stronger influence on ET in many biomes [@novickIncreasingImportanceAtmospheric2016]. 
Given the mechanisms reviewed here, we might expect that atmospheric dryness in particular skews the stress more towards the exposed canopy trees.

@@

Thus, in synthesis, warming temperatures will affect trees across the vertical gradient, but the stress will be of a different nature at different heights.
We expect that the tallest trees will be increasingly prone to hydraulic failure and damaging or lethal $T_{leaf}$'s, whereas understory trees may be increasingly prone to carbon starvation, particularly under humid conditions that make it more challenging to shed excess heat.
Both groups -- but probably disproportionately the canopy trees -- are likely to experience increasing mortality, with the modes of mortality mirroring the current tendency for canopy trees to be more prone to climate-related disturbances including drought while understory trees are more prone to competition and carbon starvation [@gora_implications_2021].
When the canopy trees die, this will open canopy gaps, as discussed in the following section.

@@

*In the understory, warming is disproportionately affecting the less thermally-adapted plant species, resulting in associated plant community thermophilization [@zellwegerForestMicroclimateDynamics2020; @duqueThermophilizationAdultJuvenile2015; @greiserHidingClimateCharacterizing2019].*
